Testing Quasar Unification: I. Radiative Transfer in Clumpy Winds

### 2015-12-18 paper.tex
added param grid
version for collabs

%% Lines and ions
\def\civ{C~\textsc{iv}}
\def\nv{N~\textsc{v}}
\def\hei{He~\textsc{i}}
\def\heii{He~\textsc{ii}}
\def\heiiline{He~\textsc{ii}~$4686$\AA}
\def\mg{Mg~\textsc{ii}}
\def\al{Al~\textsc{iii}}
\def\heii{He~\textsc{ii}}
\def\ovi{O~\textsc{vi}}
\def\la{Ly~$\alpha$}
\def\ha{H~$\alpha$}
\def\hb{H~$\beta$}

@@

\title
[Testing Quasar Unification]{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
%Unifying quasars with clumpy wind models
}

%\author[Matthews et al.]
\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$,
Nick~Higginbottom$^1$, Knox~S.~Long$^2$, Stuart~A.~Sim$^3$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

\date{\today}
%\\
%Supervisor: Prof. Christian Knigge\\
%{\sl School of Physics \& Astronomy, University of Southampton,
%  Southampton, SO17 1BJ, UK}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract} 
% Broad absorption lines (BALs) in the ultraviolet 
% are seen in $\sim20\%$ of quasi-stellar objects (QSOs). 
% Blue-shifted broad absorption lines (BALs) are the most direct evidence of 
% accretion disc `winds' in such systems; mass loaded outflows
% emanating from the disc that may be driven by line forces or
% magnetic processes. 
% Various unification schemes for
% quasars and luminous active galactic nuclei (AGN) have proposed
% that the broad emission line region is roughly cospatial
% with broad absorption line (BAL) gas and much of the phenomenology of luminous AGN
% can be explained by a simple geometrical picture involving an accretion
% disc and associated outflow. Here, we test this paradigm by 
% utilising our state-of-the-art radiative transfer code to produce synthetic spectra
% from simple biconical disc wind models. 
% In particular, we expand on our previous 
% work in which a benchmark model for BAL quasars was produced. 
% We have conducted a limited parameter search with the aim of unifying quasar phenomenology,
% and arrive at an improved model. The grid is now publicly available.
% We find that a simple treatment of clumping (`microclumping') 
% allows for a more realistic X-ray luminosity in the model by lowering the 
% ionization parameter. We examine the X-ray properties of this new model
% and find good agreement with existing X-ray samples of AGN and QSOs.
% We find that the dense, X-ray heated wind 
% produces strong H recombination and collisionally excited resonance 
% line emission to emerge at the low inclination angles, 
% which represent quasars within this unification scenario.
% % We also treat Hydrogen using a `macro-atom' approach in order to 
% % examine the effect of recombination on Hydrogen emission lines, and
% % this results in significant line emission in Lyman~$\alpha$ and the Balmer series.
% However, we are unable to reproduce the remarkably similar of line-to-continuum
% ratios between BAL and non-BAL quasar composites.
% This is due to a fundamental constraint arising
% from the anisotropy of emission from a classical thin disc. 
% We briefly explore the effect of relativistic beaming, gravitational redshift and 
% light bending on the angular distribution of disc continuum emission. 
% We find that these general relativistic effects do cause the disc to
% emit slightly more isotropically, but this is not yet sufficient to produce a
% self-consistent model. We discuss wind reprocessing as a potential solution.
% Overall, our work suggests that geometric unification
% involving an accretion disc wind is a promising scenario, but our results 
% pose a number of difficult challenges to such a model.
% Determining the true geometry of ADWs and uncovering the true disc spectral 
% (and angular) energy distribution are key next stpng if we are to build up a 
% holistic picture of the quasar population.
Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing quasar spectra to 
synthetic spectra of simple biconical disc wind models, produced with our state-of-the-art 
Monte Carlo radiative transfer and photoionization code PYTHON.  
Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars 
for plausible wind models, but only if the X-ray luminosity was limited to about  
10$^{43}$~erg~s$^{-1}$. Here, we investigate the  degree to which
clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame ultraviolet (UV) 
that have the characteristics of quasars with X-ray luminosities as 
large as 10$^{45}$ erg s$^{-1}$. 
We conclude, perhaps not surprisingly, that clumping does maintain the ionization 
state of the wind necessary for strong BAL features in the rest-frame UV at more 
realistic X-ray luminosities. We examine the X-ray properties of our fiducial clumped model
and find good agreement with existing X-ray samples of AGN and quasars.
In addition, the dense, X-ray heated wind produces strong recombination and collisionally 
excited line emission in, e.g., \la and \civ\, to emerge at the low inclination, 
`Type 1 quasar-like' angles. At the highest inclinations, the synthetic spectra 
possess prominent \mg\ and \al\ BALs, the absorption features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce all the emission lines seen in
quasar spectra with the correct equivalent width 
ratios in any of the models in our initial simulation grid,
and we find an angular dependence of emission line equivalent width despite the 
remarkable similarities in the observed spectral properties of BAL and non-BAL quasars.
Overall, our work suggests that BAL outflows can produce significant broad emission
lines and that accretion disc winds naturally reproduce the range of
ionization states seen in AGN, but a full quantitative agreement 
with quasar and BALQSO properties is not forthcoming.
% the remarkably uniform emission line properties 
% seen in BAL and non-BAL quasar composites, as we find a marked increase
% in emission line equivalent widths at high inclination `BALQSO-like' angles. 
% This is due to a fundamental constraint arising
% from the anisotropy of emission from a classical thin disc. 
% Overall, our work suggests that BAL outflows can produce significant broad emission
% lines and that accretion disc winds naturally reproduce the range of
% ionization states seen in AGN.
% This suggests that geometric unification
% involving an accretion disc wind is a promising scenario, but our results 
% pose a specific challenge to models in which an equatorial outflow
% rises from a limb-darkened, foreshortened accretion disc.
% {\bf There is confusion in the abstract, at least, in what is the model and what is the synthetic spectra, as well as whether there is one model or many models.  We need to try to be clear.  The model should be a clumped biconical outflow.  There are then particular instantiations of the model. And then there are synthetic spectra}
\end{abstract}

%\begin{keywords}
%AGN: outflows
%\end{keywords}

@@

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emission from a 
geometrically thin, optically thick accretion disc surrounding the central black hole, similar
to that described by \cite{shakurasunyaev1973}.
% {\bf Note: the first half of this paragraph and the second do not follow from one another.  
% The first half is about spectra the second about outflows and the connection to spectra is not made until the next paragrph.  
% This needs fixing. JM: Knox, I like this structure...I start with inflow, then move to outflow.}
In addition to the {\em inflowing} accreting material, 
{\em outflows} are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind. 
According to this paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows  in radio-loud (RL) sources \citep{zhou2006,ghoshpunsly2007}.
% although this geometry is contradicted
% by more recent studies of radio spectral indices \citep{dipompeo2012}.

Due to their ubiquitous nature,
disc winds offer a natural way to {\em unify} much of the
diverse phenemonology of luminous AGN and quasars \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this geometric unification framework, the broad-line region (BLR) can 
correspond either to the dense wind base or clumps embedded
in the outflow. 
% Indeed, \cite{elitzur2014} show that a disc-wind BLR scenario
% naturally explains the emission line evolution of AGN.
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit high ionization line absorption; LoBALQSOs, which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs, which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified geometric models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

% As well as acting as a source of photoionized plasma, a 
% As well as imprinting clear line absorption and emission
% features, disc winds may also have a profound effect on the structure and 
% emergent {\em continuum} of the accretion disc itself.
% Mass-loss will alter the accretion rate and resultant 
% temperature of the accretion disc, possibly explaining some 
% of the features typically seen in luminous AGN \citep{knigge1999,laordavis2014}.
% There have been numerous difficulties when confronting 
% theoretical accretion disc models with observations 
% \citep[see e.g.][]{blaes1998}. However, AGN spectral energy distributions (SEDs) can now, 
% in general, be fitted well with accretion disc models
% when the effects of general relativity (GR), Comptonisation
% and mass-loss are included \citep{capellupo2015}. Mass-loss therefore appears to be 
% critical if an accretion disc model is to successfully fit AGN SEDs, 
% particular in the UV region of the spectrum.

%{\bf ksl: I don't see what this paragraph has to do with our problem, which we need to clearly explain} Despite the clear importance of ADWs in understanding AGN SEDs and accretion physics,
Despite the clear importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
% The presence of line-locked features \citep{bowler2014} 
% and the `ghost of \la' (Arav et al. 1996; Arav 1996; North 2006; but see 
% also Cottis et al. 2010) \nocite{arav1995, arav1996, north2006,cottis2010}
% in the spectra of some BALQSOs also gives clearer evidence that line-driving is
% at least partially contributing to the acceleration of the wind.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma, meaning that it is difficult to prevent 
the wind becoming over-ionized and `failing' in the presence of strong X-rays. 
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
% Hydrodynamic simulations of line-driven disc winds also found a shielding region
% was required to maintain the correct ionization state \citep{PSK2000,PK04}. 
% However, \cite{sim2010_hydro} and \cite{H14} showed that including multiple scattering means the ionizing radiation field could still reach the previously shielded regions 
% in those particular models.
An additional or alternative solution is that the wind is clumped \citep[e.g.][]{hamann2013}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma while still maintaining the same mass-loss 
rate and column density.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
Clumping can be caused by magnetic confinement \citep{dekool1995},
or the instabilities inherent to line-driven winds 
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984,owockirybicki1985}.
Additionally, clumping is required to explain the electron scattering wings of emission lines formed
in line-driven hot star winds \citep{hillier1991eswingsmodel}. Complex substructures 
are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to line-driven instabilities 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
% {\bf KSL: I do not believe the AGN simulations have the resolution to produce clumping on the scale of hot star winds.
% JM: that's why I mention scales, although I can specifically mention resolution if we
% want.}
Nevertheless, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario.

@@

Here we attempt to address both of these issues, by introducing clumping into our model and a more complete treatment of H and He into our radiative transfer calculations.   The remainder of this paper is organized as follows:
% There has been some success using simple kinematic prescriptions for biconical disc winds to model  AGN and quasar outflows \citep[][hereafter H13]{simlong2008,sim2010,higginbottom2013}.  H13 successfully produced a benchmark model for BALQSOs. However, the model had two key drawbacks. First, an unrealistically low X-ray luminosity was required in order to prevent over-ionization of the outflow. Second, the model failed to produce the strong emission lines  required at low inclinations in a unified model. In this paper, we attempt to address these issues, and test the disc wind  unification model using Monte Carlo radiative transfer (MCRT) and photoionization  calculations. The paper is organised as follows.
In section 2, we describe some of the important photoionization 
and MCRT aspects of the code. We then outline the model in section 3, including 
a description of our clumping implementation and success criteria. 
Section 4 contains the results from a clumped model, which we discuss results
in comparison to observational data. Finally, we summarise our findings in section 5.

@@

The code operates as follows:   This outflow is discretized into $n_x \times n_z$ cells in a 2.5D
cylindrical geometry with azimuthal symmetry. From some initial conditions in each cell, the code first calculates the ionization structure of the wind in a series of iterations.  Each iteration in an ``ionization cycle'' consists of generating photons, actually photon packets, 
from an accretion disc and central object, and calculating how these photon 
bundles scatter through the wind (eventually escaping the outflow or hitting the disk). 
The ionization and temperature structure is updated based on the properties of the 
radiation field in each cell, and the process is repeated. The radiative transfer and thermal
balance is carried out according to the principle of radiative equilibrium, 
which assumes both statistical equilibrium and that radiative heating balances radiative cooling.
Once the ionization structure has converged, it is held fixed, 
and synthetic spectra are generated at specific inclination 
angles in a series of "spectral cycles". LK02 provide a more detailed  
description of the original code; various improvements have been made 
since then and are described by \cite{simmacro2005}, H13 and M15.  
%A full description of the code will be published elsewhere; 
We focus here on the specific changes made for this study
intended to improve the ionization calculation of H and He
and to allow for clumping in the wind.
%These studies contain extensive detail on the code,  so we only briefly describe the key elements of the global ionization calculation and other important aspects.

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}). 
which are reprocessed by `macro-atoms' with associated transition probabilities.
% according to radiative equilibrium, which assumes both statistical equilibrium 
% and that radiative heating balances radiative cooling.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as a macro-atom, because 
we expect recombination to be important
in determining their level populations and resultant line emission, 
and because we are especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.  
% {\bf KSL:
% I thought we were also doing He. JM: yep, corrected.}

@@

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$ in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-ions. This dilute approximation 
is not required for macro-atom levels. 
% \begin{equation}
% J_{\nu,i}=K_{pl}\nu^{\alpha_{pl}},
% \end{equation}
% for a band $i$, or an exponential 
% \begin{equation}
% J_{\nu,i}=K_{exp} e^{(-h\nu/k T_{exp})}.
% \end{equation}
% Here, $K_{exp}$, $K_{pl}$, $T_{exp}$ and $\alpha_{pl}$ are spectral fit parameters
% deduced from the band-limited radiation field estimators.
% The ionization rate out of ion $j$ can then be written as 
% \begin{equation}
% R_{j,j+1}(J)= 
% \displaystyle{
% n_j \left(C_{j} n_e + 
% \sum_{band~i=0}^{n}~
% {\int_{\nu_i}^{~\nu_i+1} \frac{4 \pi J_{\nu,i}\sigma_j(\nu)}  {h \nu} d\nu}
% \right),}
% \end{equation}
% where $\sigma_j$ is the photoionization cross-section, $n_e$
% is the electron density and $C_{j}$
% represents the collisional ionization coefficient.
% The recombination rate {\em into} ion $j$ is given by 
% \begin{equation}
% R_{j+1,j}(T_e) = (\alpha^j_{RR} + \alpha^j_{CR}) n_{j+1} n_e,
% \end{equation}
% where each $\alpha^j$ here is the recombination rate coefficient into the ground state of ion $j$.
% The subscripts denote radiative and 
% collisional (three-body) recombination, respectively. 
% $T_e$ is the electron temperature.
% %% We neglect recombination to and from excited states in the simple-ion calculation. 
% For simple-ions, we use a dilute Boltzmann approximation to calculate 
% the population of level $k$ in ionic stage $j$,
% \begin{equation}
% \frac{n_{jk}}{n_j} = \frac{W g_k}{z_j(T_R)} \exp(-E_k/kT_R).
% \end{equation}
% Here $z_j$ is the partition function of ionic stage $j$,
% $T_R$ is the effective radiation temperature,
% $E_k$ is the energy difference between level $k$ and the ground state,
% and $g_k$ is the statistical weight of level $k$. 
% We stress that this approximation is not required for ions
% treated as macro-atoms. 
% {\bf KSL: 
% It seems a bit odd that you explain what we do with simple ions our old approach, but don't say what is done for macro-atoms,.  You only say it is not the other.
% JM: We've shortened the section on simple-ions as agreed..}

@@

We include two sources of r-packets in our model:
An accretion disc and central X-ray source.
The accretion disc is assumed to be geometrically thin, but optically thick.
Accordingly, we treat the disc as an ensemble of blackbodies with a 
\cite{shakurasunyaev1973} effective temperature profile. 
The emergent SED is then determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are opaque, meaning
that r-packets that strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~r_G$, where 
$r_G = GM_{BH}/c^2$ is the gravitational radius.
For a $10^9~M_\odot$ black hole, this is equal to $8.8\times10^{14}~{\rm cm}$ 
or $\sim10^{-4}~{\rm pc}$.

@@

% is that it moderates the ionization state and temperature structure of the plasma

% Clumping the wind has an important effect on the ionization state and has
% been proposed as a solution to the so-called `over-ionization problem' in 
% disc winds \citep{hamann2013}. This is the main motivation for incorporating microclumping
% into our model. This treatment is necessarily simple; it does not adequately
% represent the complex substructures and stratifications in ionization
% state we expect in AGN outflows. Nevertheless, this parameterization allows simple estimates
% of the effect clumping might have on the ionization state and emergent 
% line emission.
% It is also encouraging that microclumping has been used 
% successfully in fits to O-star wind spectra \citep{hillier1991eswingsmodel}.

\subsection{The Simulation Grid}
% Arriving at a next-generation model}

>>>>

%% Lines and ions
\def\civ{C~\textsc{iv}}
\def\nv{N~\textsc{v}}
\def\hei{He~\textsc{i}}
\def\heii{He~\textsc{ii}}
\def\heiiline{He~\textsc{ii}~$4686$\AA}
\def\mg{Mg~\textsc{ii}}
\def\al{Al~\textsc{iii}}
\def\heii{He~\textsc{ii}}
\def\ovi{O~\textsc{vi}}
\def\la{Ly~$\alpha$}
\def\ha{H~$\alpha$}
\def\hb{H~$\beta$}
\def\civline{C~\textsc{iv}~$1550$\AA}
\def\nvline{N~\textsc{v}~$1240$\AA}
\def\mgline{Mg~\textsc{ii}~$2800$\AA}

@@

\title
[Testing Quasar Unification: Radiative Transfer in Clumpy Winds]{
% Modelling quasar outflows:
% clumpy winds and broad emission lines
Testing Quasar Unification: Radiative Transfer in Clumpy Winds
%Unifying quasars with clumpy wind models
}

\author[Matthews et al.]{
\parbox[t]{\textwidth}{
James~H.~Matthews$^1$\thanks{jm8g08@soton.ac.uk}, Christian~Knigge$^1$,
Nick~Higginbottom$^1$, Knox~S.~Long$^2$, Stuart~A.~Sim$^3$ and Sam~W.~Mangham$^1$
}
\medskip  
\\$^1$School of Physics and Astronomy, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom
\\$^2$Space Telescope Science Institute, 3700 San Martin Drive, Baltimore, MD, 21218
\\$^3$School of Mathematics and Physics, Queens University Belfast, University Road, Belfast, BT7 1NN, Northern Ireland, UK
}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%          ABSTRACT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract} 
Various unification schemes have been proposed to interpret the complex phenomenology 
of quasars and luminous active galactic nuclei (AGN) in terms of  a simple axisymmetric 
picture involving a central black hole, an accretion disc and an associated outflow.
Here, we continue our tests of  this paradigm by comparing quasar spectra to 
synthetic spectra of simple biconical disc wind models, produced with our state-of-the-art 
Monte Carlo radiative transfer and photoionization code PYTHON.  
Previously, we have shown that we could produce synthetic spectra 
resembling those of observed broad absorption line (BAL) quasars 
for plausible wind models, but only if the X-ray luminosity was limited to about  
10$^{43}$~erg~s$^{-1}$. Here, we investigate the  degree to which
clumping of the outflow allows us to 
produce synthetic spectra in the rest-frame ultraviolet (UV) 
that have the characteristics of quasars with X-ray luminosities as 
large as 10$^{45}$ erg s$^{-1}$. 
We conclude, perhaps not surprisingly, that clumping does maintain the ionization 
state of the wind necessary for strong BAL features in the rest-frame UV at more 
realistic X-ray luminosities. We examine the X-ray properties of our fiducial clumped model
and find good agreement with existing X-ray samples of AGN and quasars.
In addition, the dense, X-ray heated wind produces strong recombination and collisionally 
excited line emission in, e.g., \la\ and \civline, to emerge at the low inclination, 
`Type 1 quasar-like' angles. At the highest inclinations, the synthetic spectra 
possess prominent \mg\ and \al\ BALs, the absorption features seen in LoBAL quasars. 
Despite these successes, we are unable to reproduce all the emission lines seen in
quasar spectra with the correct equivalent width 
ratios in any of the models in our initial simulation grid,
and we find an angular dependence of emission line equivalent width despite the 
remarkable similarities in the observed spectral properties of BAL and non-BAL quasars.
Overall, our work suggests that a biconical wind model
reproduces much of the qualitative behaviour expected from a unified model. 
However, we cannot yet provide quantitative matches with quasar properties at all 
viewing angles; whether disc winds can successfully unify quasars is therefore 
still an open question.
\end{abstract}

% \begin{keywords}
% galaxies: active; accretion, accretion discs; (galaxies:) quasars: emission lines;
% (galaxies:) quasars: absorption lines
% \end{keywords}

@@

The spectra of 
quasars and luminous active galactic nuclei (AGN) 
typically exhibit a series of strong emission lines
with an underlying blue continuum - the so-called {\sl `big blue bump'} (BBB). 
The BBB is often attributed to emission from a 
geometrically thin, optically thick accretion disc surrounding the central black hole (BH), similar to that described by \cite{shakurasunyaev1973}.
% {\bf Note: the first half of this paragraph and the second do not follow from one another.  
% The first half is about spectra the second about outflows and the connection to spectra is not made until the next paragrph.  
% This needs fixing. JM: Knox, I like this structure...I start with inflow, then move to outflow.}
In addition to the {\em inflowing} accreting material, 
{\em outflows} are ubiquitous in AGN
and quasars \citep{kellerman1989,ganguly2008}. These outflows can take the form of 
highly collimated radio jets \citep[e.g.][]{hazard1963,potash1980,perley1984,marscher2006}, 
or mass-loaded `winds' emanating from the accretion disc 
\citep{weymann1991,turnermiller2009}. 
Outflows in AGN offer a 
potential feedback mechanism through which the central source can 
affect its environment \citep{king2003,king2005,fabian2012}
-- feedback that is required in models of galaxy evolution \citep{springel2005}
and may explain the `$M-\sigma$' relation \citep{silkrees1998,haring2004}.

Perhaps the clearest evidence of outflows in AGN is  
the blueshifted ($\sim 0.1c$) broad absorption lines (BALs) in the 
ultraviolet seen in approximately $20\%$ of quasars
\citep{weymann1991, reichard2003, knigge2008, turnermiller2009, allen2011}.
The simplest explanation for the incidence of 
BAL quasars (BALQSOs) is in terms of an accretion disc wind. 
According to this paradigm, a biconical wind rises from 
the accretion disc and the BALQSO fraction is associated with
the covering factor of the outflow. 
Polarisation studies expect the wind to be roughly equatorial
\citep{goodrich1995, cohen1995}, although there is also evidence
for polar BAL outflows  in radio-loud (RL) sources \citep{zhou2006,ghoshpunsly2007}.

Due to their ubiquitous nature,
disc winds offer a natural way to {\em unify} much of the
diverse phenemonology of luminous AGN and quasars \citep[e.g.][]{MCGV95, elvis2000}. 
Depending on viewing angle, an observer 
may see a BALQSO or normal `Type 1' quasar.
Within this geometric unification framework, the broad-line region (BLR) can 
correspond either to the dense wind base or clumps embedded
in the outflow. 
A biconical wind model can also readily explain the various sub-classifications of BALQSOs: 
HiBALQSOs, which only exhibit high ionization line absorption; LoBALQSOs, which also show
absorption in lower ionization state species such as \mg\ and \al; and
FeLoBALQSOs, which show further absorption in Fe~\textsc{ii} and \textsc{iii}.
In unified geometric models, this is generally attributed to ionization stratification
of the outflow \citep[e.g.][]{elvis2000}.

Despite the clear importance of disc winds in shaping quasar and AGN spectra,  
much of the underlying outflow physics remains highly uncertain. 
Several possible driving mechanisms have been proposed, including
thermal pressure \citep{weymann1982, begelman1991}, magnetocentrifugal forces 
\citep{blandfordpayne,pelletier_pudritz} and 
radiation pressure on spectral lines \citep[`line-driving';][]{lucysolomon1970,shlosman1985,MCGV95}.
Of these, line-driving is possibly the most attractive, as
strong absorption lines are already seen in BALQSOs and the X-ray spectra of AGN 
\citep{reeves2003,poundsreeves2009,tombesi2010a}.
The efficiency of line-driving is crucially dependent on the ionization state 
of the outflowing plasma, meaning that it is difficult to prevent 
the wind becoming over-ionized and `failing' in the presence of strong X-rays. 
\cite{MCGV95} proposed a potential solution: 
a region of `hitchhiking gas' that could shield the wind from the central X-ray source. 
An additional or alternative solution is that the wind is clumped 
\citep[e.g.][]{hamann2013}
possibly on multiple scale lengths. Local density enhancements could lower the 
ionization parameter of the plasma while still maintaining the same mass-loss 
rate and column density.

Evidence for dense substructures in AGN winds is widespread.
BALQSOs show complex absorption line profiles \citep{ganguly2006, simonhamann2010}
and exhibit variability in these profile shapes \citep{capellupo2011,capellupo2012,capellupo2014}.
AGN generally show variability in X-ray absorption components \citep[e.g.][]{risaliti2002}
and many models for the BLR consist of clumps embedded in an outflow 
\citep{krolik1981, emmering1992, dekool1995, cassidyraine1996}.
Clumping can be caused by magnetic confinement \citep{dekool1995},
or the instabilities inherent to line-driven winds 
\citep{lucysolomon1970,macgregor1979,carlberg1980,owockirybicki1984,owockirybicki1985}.
Additionally, clumping is required to explain the electron scattering wings of emission lines formed
in line-driven hot star winds \citep{hillier1991eswingsmodel}. Complex substructures 
are also produced in simulations of line-driven 
outflows in AGN, although on very different scales to line-driven instabilities 
\citep{PSK2000,PK04,progakurosawa2010,proga2014}.  
Nevertheless, clumpy winds offer an observationally motivated and theoretically 
predicted way to lower the ionization state of a plasma, possibly in tandem
with a shielding scenario.

@@

Here we attempt to address both of these issues, by introducing clumping into our model and a more complete treatment of H and He into our radiative transfer calculations.   The remainder of this paper is organized as follows:
In section 2, we describe some of the important photoionization 
and MCRT aspects of the code. We then outline the model in section 3, including 
a description of our clumping implementation and success criteria. 
Section 4 contains the results from a clumped model, which we discuss results
in comparison to observational data. Finally, we summarise our findings in section 5.

@@

The code operates as follows:   This outflow is discretized into $n_x \times n_z$ cells in a 2.5D
cylindrical geometry with azimuthal symmetry. From some initial conditions in each cell, the code first calculates the ionization structure of the wind in a series of iterations.  Each iteration in an ``ionization cycle'' consists of generating photons, actually photon packets, 
from an accretion disc and central object, and calculating how these photon 
bundles scatter through the wind (eventually escaping the outflow or hitting the disk). 
The ionization and temperature structure is updated based on the properties of the 
radiation field in each cell, and the process is repeated. The radiative transfer and thermal
balance is carried out according to the principle of radiative equilibrium, 
which assumes both statistical equilibrium and that radiative heating balances radiative cooling.
Once the ionization structure has converged, it is held fixed, 
and synthetic spectra are generated at specific inclination 
angles in a series of "spectral cycles". LK02 provide a more detailed  
description of the original code; various improvements have been made 
since then and are described by \cite{simmacro2005}, H13 and M15.  
We focus here on the specific changes made for this study
intended to improve the ionization calculation of H and He
and to allow for clumping in the wind.

@@

Our approach to line transfer is based upon the macro-atom implementation developed by 
\cite{lucy2002, lucy2003}, in which the energy flows through the system are described in 
terms of indivisible energy quanta of radiant or kinetic energy 
(`$r-$packets' and `$k-$packets' respectively; see also section~\ref{sec:photon_sources}). 
which are reprocessed by `macro-atoms' with associated transition probabilities.
In our case, for reasons of computational efficiency, we adopt the  hybrid macro-atom scheme 
described by M15. In this scheme, the energy packets interact with either two-level 
`simple atoms' or full macro-atoms. 
This allows one to treat non-LTE line transfer in radiative equilibrium
without approximation for elements that are identified as 
full macro-atoms, while maintaining the fast `two-level' 
treatment of resonance lines when elements are identified 
as simple atoms (see M15). In this study,
only H and He are treated as a macro-atom, because 
we expect recombination to be important
in determining their level populations and resultant line emission, 
and because we are especially interested in the contribution to 
AGN spectra of \LA.  H13 treated all atoms in a two-level approximation.

@@

In order to calculate the photoionization rate, 
we model the SED in a grid cell using the technique described by H13. In this scheme,
the mean intensity, $J_{\nu}$ in a series of $n$ bands is modeled as either a power law or exponential
in frequency $\nu$, with the fit parameters deduced from band-limited radiation field estimators.
This allows the calculation of a photoionization rate estimator. Ion abundances are
then calculated by solving the rate equations between ions. We include collisional ionization and photoionization balanced with radiative, 
dielectronic and collisional (three-body) recombination.
As in M15, we use a dilute Boltzmann approximation to calculate 
the populations of levels for simple-ions. This dilute approximation 
is not required for macro-atom levels.

@@

We include two sources of r-packets in our model:
An accretion disc and central X-ray source.
The accretion disc is assumed to be geometrically thin, but optically thick.
Accordingly, we treat the disc as an ensemble of blackbodies with a 
\cite{shakurasunyaev1973} effective temperature profile. 
The emergent SED is then determined by the specified accretion rate ($\dot{m}$)
and central BH mass ($M_{BH}$).
All photon sources in our model are opaque, meaning
that r-packets that strike them are destroyed.
The inner radius of the disc extends to the innermost 
stable circular orbit (ISCO) of the BH. 
We assume a Schwarzchild BH with an ISCO at $6~r_G$, where 
$r_G = GM_{BH}/c^2$ is the gravitational radius.
For a $10^9~M_\odot$ BH, this is equal to $8.8\times10^{14}~{\rm cm}$ 
or $\sim10^{-4}~{\rm pc}$.

@@

\subsection{The Simulation Grid}
